---
title: "RxNorm & ATC lookup"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval    = FALSE
)
```

```{r setup}
library(adRpheno)
```

Medication data in clinical research usually starts as messy free text — `"atorvastatin 40 mg"`, `"METFORMIN HCL"`, `"hctz 25mg po qd"`. Turning that into something analyzable means two API round trips:

1. **Free text → RxCUI**. The National Library of Medicine's [RxNorm](https://www.nlm.nih.gov/research/umls/rxnorm/) assigns each clinical drug a unique numeric ID (RxCUI). `add_rxcuis()` resolves names to RxCUIs.
2. **RxCUI → ATC class**. The WHO ATC classification organizes drugs into a four-level therapeutic hierarchy. `add_atc_classification()` retrieves the class at the level you ask for.

Both functions hit the live [RxNav REST API](https://rxnav.nlm.nih.gov/), so this article is website-only (excluded from `R CMD check`).

## Step 0 — Clean before you lookup

Match rates depend heavily on input quality. The API works best on **ingredient names** without dosages, formulations, or brand-name decoration.

A minimal cleaning pass:

```{r}
meds_raw <- data.frame(
  med = c("Atorvastatin 40 MG Oral Tablet",
          "METFORMIN HCL 500MG",
          "lisinopril",
          "hctz",
          "Aspirin EC 81mg")
)

# Lowercase, strip dosage/units, drop common formulation words
meds_raw$med_clean <- meds_raw$med |>
  tolower() |>
  gsub("\\d+\\s*(mg|mcg|g|ml|iu)\\b", "", x = _) |>
  gsub("\\b(hcl|tablet|oral|po|ec|cr|er|xr)\\b", "", x = _) |>
  trimws()

meds_raw
#>                              med            med_clean
#> 1 Atorvastatin 40 MG Oral Tablet         atorvastatin
#> 2            METFORMIN HCL 500MG             metformin
#> 3                     lisinopril           lisinopril
#> 4                           hctz                 hctz
#> 5                Aspirin EC 81mg              aspirin
```

You don't need anything fancier than that to get started. Iterate based on what fails to match.

## Step 1 — Look up RxCUIs

```{r}
meds <- add_rxcuis(meds_raw, med_column = "med_clean")
```

What this does under the hood:

- **De-duplicates** before calling the API. Only unique non-empty values are queried, so a column with 10,000 rows but 200 unique drug names hits the API 200 times.
- **Searches** with RxNorm's `search=2` mode (exact match, then normalized match). This is what lets `"hctz"` resolve to hydrochlorothiazide without false positives.
- **Caches** every result to `cache/rxcui_cache.rds`. Reruns are essentially free.
- **Batches** with retry on failure (3 attempts, exponential backoff: 0.5s, 1s, 2s).
- **Reports** match rate when done: `RxCUI: 5/5 (100%) in 0.1 min`.

Result:

```{r}
meds
#>                              med            med_clean    rxcui
#> 1 Atorvastatin 40 MG Oral Tablet         atorvastatin    83367
#> 2            METFORMIN HCL 500MG             metformin     6809
#> 3                     lisinopril           lisinopril    29046
#> 4                           hctz                 hctz     5487
#> 5                Aspirin EC 81mg              aspirin     1191
```

### When a match fails

Unmatched rows get `NA`. The most common causes:

- **Brand names** the cleaner didn't strip. `"glucophage"` won't match; `"metformin"` will. Either map brand to generic up front, or use [`get_single_rxcui()`](../reference/get_single_rxcui.html) interactively to test problem cases.
- **Compound formulations**. `"lisinopril/hctz"` is two drugs; RxNorm wants them split into separate rows first.
- **Multi-word names with extra qualifiers**. `"vitamin d3"` may match where `"vitamin d 3 (cholecalciferol)"` doesn't.

For a single tricky name, debug interactively without touching the batch cache:

```{r}
get_single_rxcui("glucophage")
#> [1] "6809"  (yes, the API does brand-name lookup directly)

get_single_rxcui("lisinopril/hctz")
#> [1] NA      (compound; split it)
```

## Step 2 — Attach ATC classifications

Once every row has an RxCUI, ask for ATC classes:

```{r}
meds <- add_atc_classification(meds, rxcui_col = "rxcui", atc_level = "second")
meds
#>                              med            med_clean    rxcui                    atc2_class
#> 1 Atorvastatin 40 MG Oral Tablet         atorvastatin    83367         lipid modifying agents
#> 2            METFORMIN HCL 500MG             metformin     6809         drugs used in diabetes
#> 3                     lisinopril           lisinopril    29046  agents acting on the renin-...
#> 4                           hctz                 hctz     5487                      diuretics
#> 5                Aspirin EC 81mg              aspirin     1191         antithrombotic agents
```

### Choosing an ATC level

ATC is a four-level hierarchy:

| Level    | Width   | Example  | Meaning                       |
|----------|---------|----------|-------------------------------|
| `first`  | 1 char  | `C`      | Anatomical main group         |
| `second` | 3 chars | `C10`    | Therapeutic subgroup          |
| `third`  | 4 chars | `C10A`   | Pharmacological subgroup      |
| `fourth` | 5 chars | `C10AA`  | Chemical subgroup             |

`"second"` is the most common choice for analysis grouping — granular enough to distinguish "lipid modifying" from "antithrombotic," coarse enough that you get a manageable number of categories. Use `"third"` or `"fourth"` if you need finer distinctions like statins vs. PCSK9 inhibitors.

Each level uses a separate cache file (`atc1_cache.rds`, `atc2_cache.rds`, ...) so switching levels doesn't invalidate previous work.

### Drugs with multiple classifications

Some drugs sit in more than one ATC class — aspirin is both `antithrombotic agents` and `analgesics`, for example. By default the result is collapsed:

```{r}
meds$atc2_class[5]
#> [1] "antithrombotic agents; analgesics"
```

For tidyverse-style downstream work where you'd rather have one row per (drug, class), use a list-column:

```{r}
meds_list <- add_atc_classification(
  meds_raw_with_rxcui,
  rxcui_col      = "rxcui",
  as_list_column = TRUE
)
# Then unnest if you want one row per (drug, class):
tidyr::unnest_longer(meds_list, atc2_class)
```

## Tuning for larger datasets

The defaults are tuned for a few thousand unique drugs. For larger jobs the parameters worth knowing:

| Parameter      | Default | What it controls                                  |
|----------------|---------|---------------------------------------------------|
| `batch_size`   | 500     | API calls per batch before the inter-batch sleep  |
| `batch_delay`  | 0.5     | Seconds between batches (rate-limit hygiene)      |
| `save_freq`    | 500     | How often the cache is written to disk            |
| `retry_count`  | 3       | Retry attempts on failed calls                    |
| `max_age_days` | 30      | Cache entry expiration                            |

A typical "big job" tuning — say 50,000 unique drug strings — bumps batch sizes up and saves more often so an interruption doesn't lose much progress:

```{r}
add_rxcuis(meds_raw,
           med_column   = "med_clean",
           batch_size   = 1000,
           save_freq    = 200,
           batch_delay  = 1)
```

If you're being rate-limited (failures clustered in time, not random), increase `batch_delay` rather than decreasing `batch_size` — that's the lever RxNav cares about.

## What comes next

Once medications have RxCUIs and ATC classes, you usually want to either categorize them by therapeutic class (statins, antidiabetics, ...) or pivot from one-row-per-medication to one-row-per-participant. That's [the next article](medications-categorize-pivot.html).
